Untargeted control doc

### 2023-09-26 untargeted/paruntargeted_control.Rmd
Attempt at automating annotation before deciding to switch to manual

```{r making peak_data and feat_data}
msnexp_filled <- readRDS(paste0(output_folder, "msnexp_filled.rds"))
peak_data_long <- msnexp_filled %>%
  chromPeaks() %>%
  as.data.frame() %>%
  rownames_to_column() %>%
  mutate(peakidx=row_number())
peak_data <- msnexp_filled %>%
  featureDefinitions() %>%
  as.data.frame() %>%
  select(mzmed, rtmed, npeaks, peakidx) %>%
  rownames_to_column("id") %>%
  unnest_longer(peakidx) %>%
  rename_with(~paste0("feat_", .x), .cols = -peakidx) %>%
  dplyr::rename(feature="feat_id") %>%
  left_join(peak_data_long, by = join_by(peakidx)) %>%
  mutate(filename=basename(fileNames(msnexp_filled))[sample])
output_csv_write(peak_data)

@@

best_feat_data <- all_feat_data %>%
  filter(med_cor>0.75) %>%
  select(-med_cor, -med_snr)
output_csv_write(best_feat_data)
```

```{r diagnostic plots}
mass_check <- function(mz_i){
  mass_feats <- best_feat_data %>% 
    filter(mzmed%between%pmppm(mz_i, 5)) %>%
    print()
  chosen_eic <- open_dataset("tmzMLs/pqds") %>%
    filter(polarity==!!polarity) %>%
    filter(amendment==!!amendment) %>%
    filter(mz%between%pmppm(mz_i, 5)) %>%
    filter(rt%between%c(2, 20)) %>%
    dplyr::collect()
  ggtitle_i <- ifelse(is.null(names(mz_i)), mz_i, names(mz_i))
  chosen_eic %>%
    left_join(parametadata, by = join_by(filename, samp_type, amendment, polarity)) %>%
    mutate(samp_type=str_extract(filename, "Mix\\d|Deep|Surf|Poo|Blk")) %>%
    filter(!str_detect(filename, "Mat")) %>%
    mutate(plot_facet=ifelse(samp_type%in%c("Mix1", "Mix2"), "Mixes", "Samples")) %>%
    ggplot() +
    geom_line(aes(x=rt, y=int, group=filename, color=samp_type)) +
    facet_wrap(~plot_facet, scales = "free_y", ncol=1) +
    geom_vline(aes(xintercept=rtmed), color="black", data = mass_feats) +
    ggtitle(ggtitle_i) +
    theme_bw()
}
mass_check(c("Carn/THB/HMSB"=162.1125))
mass_check(c("ACH/TMAB"=146.1181))
mass_check(c("Ala/bAla/Sarc"=90.0555))
mass_check(c("Arg"=175.119501))
mass_check(c("Leu/Ile/TMAP"=132.1025))
mass_check(c("15N Ile"=132.1025+0.997))
```

>>>>

```{r making peak_data and best_feat_data}
msnexp_filled <- readRDS(paste0(output_folder, "msnexp_filled.rds"))
peak_data_long <- msnexp_filled %>%
  chromPeaks() %>%
  as.data.frame() %>%
  rownames_to_column() %>%
  mutate(peakidx=row_number())
peak_data <- msnexp_filled %>%
  featureDefinitions() %>%
  as.data.frame() %>%
  select(mzmed, rtmed, npeaks, peakidx) %>%
  rownames_to_column("id") %>%
  unnest_longer(peakidx) %>%
  rename_with(~paste0("feat_", .x), .cols = -peakidx) %>%
  dplyr::rename(feature="feat_id") %>%
  left_join(peak_data_long, by = join_by(peakidx)) %>%
  mutate(filename=basename(fileNames(msnexp_filled))[sample])
output_csv_write(peak_data)

@@

if(amendment=="Amm"){
  if(polarity=="pos"){
    rescue_feats <- c("FT3015", "FT1319", "FT1321")
  } else if(polarity=="neg"){

  }
}
if(amendment=="Nit"){
  if(polarity=="pos"){

  } else if(polarity=="neg"){

  } else {
    stop("Unknown polarity")
  }
}
if(amendment=="GMP"){
  if(polarity=="pos"){

  } else if(polarity=="neg"){

  } else {
    stop("Unknown polarity")
  }
}
if(amendment=="Arg"){
  if(polarity=="pos"){

  } else if(polarity=="neg"){

  } else {
    stop("Unknown polarity")
  }
}

best_feat_data <- all_feat_data %>%
  filter(med_cor>0.75 | feature%in%rescue_feats) %>%
  select(-med_cor, -med_snr)
output_csv_write(best_feat_data)
```

```{r diagnostic plots}
mass_check <- function(mz_i, all_samps=FALSE){
  mass_feats <- best_feat_data %>% 
    filter(mzmed%between%pmppm(mz_i, 5)) %>%
    print()
  if(all_samps){
    chosen_eic <- open_dataset("tmzMLs/pqds") %>%
      filter(polarity==!!polarity) %>%
      filter(amendment==!!amendment) %>%
      filter(mz%between%pmppm(mz_i, 5)) %>%
      filter(rt%between%c(2, 20)) %>%
      dplyr::collect()
  } else {
    chosen_eic <- open_dataset("tmzMLs/pqds") %>%
      filter(polarity==!!polarity) %>%
      filter(amendment==!!amendment) %>%
      filter(samp_type=="Poo" | samp_type=="Std") %>%
      filter(mz%between%pmppm(mz_i, 5)) %>%
      filter(rt%between%c(2, 20)) %>%
      dplyr::collect()
  }
  ggtitle_i <- ifelse(is.null(names(mz_i)), mz_i, names(mz_i))
  chosen_eic %>%
    left_join(parametadata, by = join_by(filename, samp_type, amendment, polarity)) %>%
    mutate(samp_type=str_extract(filename, "Mix\\d|Deep|Surf|Poo|Blk")) %>%
    filter(!str_detect(filename, "Mat")) %>%
    mutate(plot_facet=ifelse(samp_type%in%c("Mix1", "Mix2"), "Mixes", "Samples")) %>%
    ggplot() +
    geom_line(aes(x=rt, y=int, group=filename, color=samp_type)) +
    facet_wrap(~plot_facet, scales = "free_y", ncol=1) +
    geom_vline(aes(xintercept=rtmed), color="black", data = mass_feats) +
    ggtitle(ggtitle_i) +
    theme_bw()
}
mass_check(c("Carn/THB/HMSB"=162.1125))
mass_check(c("ACH/TMAB"=146.1181), all_samps = TRUE)
mass_check(c("Ala/bAla/Sarc"=90.0555))
mass_check(c("Arg"=175.119501))
mass_check(c("Leu/Ile/TMAP"=132.1025))
mass_check(c("15N Ile"=132.1025+0.997))
mass_check(60.04510)
mass_check(c("Ectoine"=143.0821))
mass_check(c("Betonicine"=160.0974))
mass_check(c("Adenine"=136.0623))
mass_check(c("Pyroglut"=130.0504))
mass_check(c("Glutamine"=147.076968))
mass_check(c("Cytosine"=244.093347))
mass_check(c("Proline"=116.0712))
```

```{r annotate}
all_stans <- read_csv("metadata/all_stans.csv") %>%
  filter(polarity==!!polarity)

stan_mix_data <- peak_data %>% 
  select(feature, filename, into) %>%
  left_join(parametadata) %>%
  filter(samp_type=="Std") %>%
  filter(!str_detect(filename, "Mat")) %>%
  mutate(mix=str_extract(filename, "Mix\\d")) %>%
  select(feature, mix, into) %>%
  pivot_wider(names_from = mix, values_from = into, values_fn = mean)

stan_annotations <- best_feat_data %>%
  pmap(function(...){
    row_data <- list(...)
    all_stans %>%
      select(compound_name, mz, rt) %>%
      filter(mz%between%pmppm(row_data$mzmed, 10)) %>%
      mutate(feature=row_data$feature, mzmed=row_data$mzmed,
             rtmed=row_data$rtmed, n=row_data$n)
  }, .progress = TRUE) %>%
  bind_rows() %>%
  arrange(mz, rt, rtmed) %>%
  left_join(stan_mix_data) %>%
  left_join(all_stans %>% select(compound_name, mix)) %>%
  mutate(mix_ratio=ifelse(mix=="Mix1", Mix1/Mix2, Mix2/Mix1)) %>%
  filter(mix_ratio>3 | is.na(mix_ratio)) %>%
  # select(feature, compound_name)
  ungroup()

missed_stans <- all_stans %>%
  anti_join(stan_annotations %>% select(compound_name)) %>%
  print(n=Inf)

stan_choose <- function(df, compound_name_i, feature_i){
  anti_df <- df %>%
    filter(compound_name==compound_name_i) %>%
    filter(feature!=feature_i)
  join_vec <- join_by(compound_name, mz, rt, feature, mzmed, rtmed, n, 
                      Mix1, Mix2, mix, mix_ratio)
  df %>% anti_join(anti_df, by = join_vec)
}
stan_add <- function(df, compound_name_i, feature_i){
  add_row(df, compound_name=compound_name_i, feature=feature_i)
}
stan_remove <- function(df, compound_name_i){
  df %>% filter(!compound_name%in%compound_name_i)
}
if(amendment=="Amm"){
  if(polarity=="pos"){
    stan_annotations %>%
      stan_choose("Sarcosine", "FT0348") %>%
      stan_choose("beta-Alanine", "FT0347") %>%
      stan_choose("Cytosine", "FT0769") %>%
      stan_add("Cytosine from cytidine", "FT0770") %>%
      stan_choose("L-Proline", "FT0862") %>%
      stan_remove("5-Oxoproline") %>%
      stan_choose("L-Leucine", "FT1244") %>%
      stan_choose("L-Isoleucine", "FT1248") %>%
      stan_choose("Proline betaine", "FT1572") %>%
      stan_choose("Betonicine", "FT1842") %>%
      stan_choose("Carnitine", "FT1857") %>%
      stan_remove("Dimethylglycine") %>%
      stan_remove("4-Aminobutyric acid") %>%
      stan_remove("Creatinine from creatine?") %>%
      stan_remove("L-Valine") %>%
      stan_remove(c("L-Threonine", "L-Homoserine")) %>%
      stan_add("Threonine/Homoserine", "FT0953") %>%
      # stan_remove("Allopurinol") %>%
      stan_remove("Guanine from guanosine?") %>%
      stan_add("L-Glutamic acid", "FT1635") %>%
      group_by(feature) %>%
      group_by(compound_name) %>%
      # filter(n()>1)
      ungroup() -> stan_annotations
  } else if(polarity=="neg"){

  } else {
    stop("Unknown polarity")
  }
}
if(amendment=="Nit"){
  if(polarity=="pos"){

  } else if(polarity=="neg"){

  } else {
    stop("Unknown polarity")
  }
}
if(amendment=="GMP"){
  if(polarity=="pos"){

  } else if(polarity=="neg"){

  } else {
    stop("Unknown polarity")
  }
}
if(amendment=="Arg"){
  if(polarity=="pos"){

  } else if(polarity=="neg"){

  } else {
    stop("Unknown polarity")
  }
}

mass_check(c("Cytidine"=244.0933))
```
